---
title: 'Lab 19: Interaction in Multiple Regression'
subtitle: 'MTH 361: Probability and Statistics in the Health Sciences'
date: "Last updated: `r Sys.Date()`"
output: 
  pdf_document
header-includes:
- | 
  ```{=latex}
  \usepackage[default]{sourcesanspro}
  \usepackage[T1]{fontenc}
  ```
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, fig.width=6, fig.height=3)
```

This lab introduces the concept of a statistical interaction, specifically in the case of an interaction between a categorical predictor and a numerical predictor.

# Introduction

An important implicit assumption in the multiple regression model

\[y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p + \epsilon \]

is that when one of the predictor variables $x_j$ changes by 1 unit and the values of the other variables remain constant, the predicted response changes by $\beta_j$, regardless of the values of the other variables.

A statistical **interaction** occurs when this is assumption is not true, such that the effect of one explanatory variable $x_j$ with the response depends on the particular value(s) of one or more other explanatory variables.

This course specifically examines interaction in a two-variable setting, where one of the predictors is categorical and the other is numerical. Interaction effects between two numerical variables and between more than two variables can be complicated to interpret. A more complete treatment of interaction is best left to a specialized statistical modeling course.

This lab introduces the concept of interaction using a sample from the NHANES data.\footnote{The NHANES data were introduced in several earlier videos and labs. We can assume the data represents a simple random sample from the American population.} 

\newpage

# Interaction with NHANES

The NHANES collected information about various demographic and health variables for each participant, including total cholesterol level in mmol/L (\texttt{TotChol}), age in years (\texttt{Age}), and diabetes status (\texttt{Diabetes}, coded as either \texttt{No} or \texttt{Yes}).

```{r}
library(oibiostat)
data(nhanes.samp.adult.500)
```

The following set of questions step through exploring the association of total cholesterol with age and diabetes status, using \texttt{nhanes.samp.adult.500}, a sample of $n = 500$ adults from the larger NHANES dataset.

1. The following output shows the linear model for predicting total cholesterol level from age and diabetes status. 

    a) Write the model equation in terms of the variable names.
    
    \vfill
    
    b) Interpret the coefficients of the model, including the intercept.
    
    \vfill
    
    c) Make predictions.
    
        i. How does the predicted mean total cholesterol for a 60-year-old individual compare to that of a 50-year-old individual, if both are diabetic?
        
        \vfill
        
        ii. How does the predicted mean total cholesterol for a 60-year-old individual compare to that of a 50-year-old individual, if both are not diabetic?
        
        \vfill
        
    d) Based on the model equation from part a), write two separate model equations: one for diabetic individuals and one for non-diabetic individuals. 
    
    \vfill
    
    e) The scatterplot provided shows total cholesterol versus age. Sketch the models from part d) on your plot. Describe what you see; compare the models.

    \vfill
    
\pagebreak

```{r}
model <- lm(TotChol ~ Age + Diabetes, data=nhanes.samp.adult.500)
summary(model)
```

\vfill

```{r}
library(tidyverse)
library(colorblindr)
nhanes.samp.adult.500 %>% 
  ggplot(aes(x=Age, y=TotChol)) + 
  geom_point(aes(col=Diabetes)) + 
  facet_wrap(~Diabetes) + 
  scale_color_OkabeIto() + 
  geom_smooth(method='lm')
```

\pagebreak


A model that assumes the relationship between total cholesterol and age does not depend on diabetes status might be overly simple and potentially misleading. 

2. To explore this visually, fit two separate models for the relationship between total and cholesterol and age. R output for the models can be found on the next page.

    a) Write the fitted model predicting total cholesterol from age in diabetic individuals. Add the linear model to the scatterplot for diabetic individuals below.
    
    \vfill
    
    b) Write the fitted model predicting total cholesterol from age in non-diabetic individuals. Add the linear model to the scatterplot for non-diabetic individuals below.
    
    \vfill
    
    c) Describe what you see in the plots. Does the association between total cholesterol level and age seem different between diabetics and non-diabetics?
    
    \vfill

```{r, echo=FALSE}
nhanes.samp.adult.500 %>% 
  ggplot(aes(x=Age, y=TotChol)) + 
  geom_point(aes(col=Diabetes)) + 
  facet_wrap(~Diabetes) + 
  scale_color_OkabeIto() + 
  geom_smooth(method='lm')
```
    
\pagebreak

```{r}
Diabetes <- nhanes.samp.adult.500 %>%
  filter(Diabetes == 'Yes')

NoDiabetes <- nhanes.samp.adult.500 %>%
  filter(Diabetes == 'No')

model_yes <- lm(TotChol ~ Age, data=Diabetes)
summary(model_yes)

model_no <- lm(TotChol ~ Age, data=NoDiabetes)
summary(model_no)
```

\newpage

With the addition of another parameter (commonly referred to as an interaction term), a linear regression model can be extended to allow the relationship of one explanatory variable with the response to vary based on the values of other variables in the model. Consider the model

\[y = \beta_0 + \beta_1(Age) + \beta_2(Diabetes) + \beta_3(Diabetes \times Age). \]

The term $(Diabetes \times Age)$ is the interaction term between diabetes status and age, and $\beta_3$ is the coefficient of the interaction term. Diabetes status and age, the main independent variables in the model, are sometimes referred to as "main effect variables" in the context of a model with an interaction term.
    
3. The code on the next page fits a model for predicting total cholesterol that includes age, diabetes, and the interaction term between age and diabetes status.

    a) Write prediction equations.
    
        i. Write the overall model equation.
        
        \vfill
        
        ii. Write the model equation for diabetics.
        
        \vfill
        
        iii. Write the model equation for non-diabetics.
        
        \vfill
        
    b) Interpret the model coefficients (of the overall equation), including the interaction term.
    
    \vfill
    
    c) Make predictions.
    
        i. How does the predicted mean total cholesterol for a 60-year-old individual compare to that of a 50-year-old individual, if both are diabetic?
        
        \vfill
        
        ii. How does the predicted mean total cholesterol for a 60-year-old individual compare to that of a 50-year-old individual, if both are not diabetic?
        
        \vfill
        
        iii. Compare the predictions made in parts i. and ii. to those made in Question 1 using the model without an interaction term. How does fitting an interaction term change the model?
        
        \vfill


\pagebreak

```{r}
model_int <- lm(TotChol ~ Age + Diabetes + Age:Diabetes, data=nhanes.samp.adult.500)
summary(model_int)
```

\vfill

4. What might explain a positive association between age and cholesterol for non-diabetics, but a negative association between age and cholesterol for diabetics?

\vfill

\pagebreak

# Problem(s) to submit

On your own, complete the following problems. You may use your notes, this lab activity, or any other resources. Submit your responses in BlueLine when you are finished.

## Interaction with PREVEND

The following set of questions step through taking a closer look at the association of RFFT score with age and statin with \texttt{prevend.samp}, a sample of $n = 500$ individuals from the PREVEND data.

```{r}
data(prevend.samp)
```

```{r, echo=FALSE}
prevend.samp$Statin <- as.factor(prevend.samp$Statin)
```

The output on the next page comes from a model for predicting RFFT score from age, statin use, and the interaction term between age and statin use (`1` if used, `0` if not).

a) Write prediction equations.
    
    i. Write the overall model equation.
        
    \vfill
        
    ii. Write the model equation for statin users.
        
    \vfill
        
    iii. Write the model equation for statin non-users.
        
    \vfill
        
b) Interpret the model coefficients.
    
\vfill
    
c) Make predictions.
    
    i. How does the predicted mean RFFT score for a 55-year-old individual compare to that of a 65-year-old individual, if both are using statins?
        
    \vfill
        
    ii. How does the predicted mean RFFT score for a 55-year-old individual compare to that of a 65-year-old individual, if both are not using statins?
        
    \vfill
        
    iii. How does the predicted mean RFFT score for a 70-year-old individual using statins compare to that of a 50-year-old individual not using statins?
        
    \vfill
        
d) Is there evidence of a statistically significant interaction between age and statin use? Support your response using the model summary output __and__ the scatterplots.
    
\vfill
    
\pagebreak

```{r}
model <- lm(RFFT ~ Age + Statin + Age:Statin, data=prevend.samp)
summary(model)
```

\vfill

```{r}
library(tidyverse)
prevend.samp %>% 
  ggplot(aes(x=Age, y=RFFT)) + 
  geom_point(aes(col=Statin)) + 
  facet_wrap(~Statin) + 
  scale_color_OkabeIto() + 
  geom_smooth(method='lm')
```