---
title: "My answers"
author: "My name"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---


This week's labs will be a little different. We'll work together as a class to explore several data sets by calculating data visualizations and summary statistics using R. Follow along, and run the R code in your RMarkdown document when instructed. 

> Below each question, you'll see a right carrot like this that says "Write your response here." That's where you'll write an answer or take notes on our in-class discussion. Make sure there is an empty line above and below this carrot. (If the text turns green, you're probably doing okay.) You don't need to leave in the "Write your response here." 

Periodically, you should "Knit" your work to save a copy.

When you're finished, upload the knitted document with your responses to BlueLine.

```{r}
library(tidyverse)
library(mosaic)
```

# Exploration 1: Behavioral Risk Factor Surveillance System

The Behavioral Risk Factor Surveillance System (BRFSS) is an annual telephone survey of 350,000 people in the United States. The BRFSS is designed to identify risk factors in the adult population and report emerging health trends. For example, respondents are asked about their diet, weekly exercise, possible tobacco use, and healthcare coverage. 

Use the following command to download the dataset `cdc` from a URL. This dataset is a sample of 20,000 people from the survey conducted in 2000, and contains responses from a subset of the questions asked on the survey.

```{r}
source("http://www.openintro.org/stat/data/cdc.R")
```

Take a look at the Environment tab, where `cdc` should now be visible. Click the blue button next to the dataset name to view a summary of the 9 variables contained in the data matrix. To view the dataset itself, click on the name of the dataset; alternatively, enter the command


\vfill

Each row of the data matrix represents a single observation and each column represents a variable. Each variable corresponds to a question that was asked in the survey. For `genhlth`, respondents were asked to evaluate their general health as either "excellent", "very good", "good", "fair", or "poor". The variables `exerany`, `hlthplan`, and `smoke100` are binary variables, with responses recorded as either 0 for "no" and 1 for "yes": whether the respondent exercised in the past month, has health coverage, or has smoked at least 100 cigarettes in their lifetime. The other variables record the respondents' height in inches, weight in pounds, their desired weight (`wtdesire`), age in years, and gender.

\pagebreak

1. Complete the code below to make a scatterplot of height and weight. Once you've replaced the blank spaces with the appropriate variables, remove the `eval = FALSE` part of the R code chunk declaration. Do height and weight appear to be associated?

> Write your response here.


\vfill

The conversion from inches to meters is 1 in = 0.0254 m. Converting variable units or calculating a new variable based on another one can be done with the `mutate()` function. For example, the code below creates a new variable `height.m` that records height in meters, then saves the result in the original data set.

```{r}
cdc <- cdc %>%
  mutate(height.m = height*0.0254)
```

2.  The conversion from pounds to kilograms is 1 lb = .454 kg. Create a new variable `weight.kg` that records weight in kilograms.

> Write your response here.


\vfill

3. BMI is calculated as weight in kilograms divided by height squared in meters squared. Create a new variable `bmi` and make a scatterplot of `height` and `BMI`. Do `height` and `BMI` seem to be associated?

> Write your response here.


\pagebreak

4. A BMI of 30 or above is considered obese. Why might health agencies choose to use BMI as a measure of obesity, rather than weight?

> Write your response here.

\vfill

# Exploration 2: National Health and Nutrition Examination Survey

The National Health and Nutrition Examination Survey (NHANES) is a survey conducted annually by the US National Center for Health Statistics (NCHS). While the original data uses a survey design that oversamples certain subpopulations, the data have been reweighted to undo oversampling effects and can be treated as if it were a simple random sample from the American population.

The following questions will be explored with the NHANES data:

  1. At what age do Americans seem to reach full adult height?
  2. What proportion of Americans age 25 or older have a college degree?
  3. What is the relationship between education level and income?
  4. How much more likely is it that someone \textit{not} physically active has diabetes, compared to someone who is active?

The NHANES data are available from the NHANES package. To view the complete list of study variables and their descriptions, access the NHANES documentation page with `?NHANES`.

```{r}
library(NHANES)
data(NHANES)
```

\vfill

For convenience, descriptions of the variables used in this lab exercise are included below.
  
- `Age`: age in years at screening. Subjects 80 years or older were recorded as 80 years of age.
- `Education`: highest educational level of study participant, reported for participants aged 20 years or older. Recorded as either "8th Grade", "9 - 11th Grade", "High School", "Some College", or "College Grad".
- `Poverty`: a ratio of family income to poverty guidelines. Smaller numbers indicate more poverty; i.e., a number below 1 indicates income below the poverty level.
- `Weight`: weight, measured in kilograms.
- `Height`: standing height, measured in centimeters.
- `Diabetes`: "Yes" if the participant was told by a health professional that they have diabetes, "No" otherwise.
- `PhysActive`: coded "Yes" if the participant does moderate or vigorous-intensity sports, fitness, or recreational activities; "No" otherwise. Reported for participants 12 years or older.

\pagebreak

1. Fill in the blanks to make a plot of the ages of NHANES participants. Describe in words the distribution of ages for the study participants.

> Write your response here.

  
\vfill

2. Use the `favstats` function from the `mosaic` package to calculate summary statistics for age. Describe the distribution of age using the summary statistics. You should refer to the center and spread of the distribution in your answer.  

> Write your response here.


\vfill

3. Complete the code below to make a frequency table of physical activity. Calculate the proportion of all respondents in the NHANES data set who are physically active. Calcualte the proportion who are not physically active.

> Write your response here.


\vfill

\pagebreak

4. Physical activity may be related to age. Complete the code below to make a bar chart of physical activity. The `filter(Age < 25)` filters the data set to those under age 25. Describe the bar chart.

> Write your response here.


\vfill

5. Complete the code below to make a bar chart of physical activity for those age 25 or older. Describe the bar chart. How does physical activity compare for the two age groups?

> Write your response here.


\vfill

\pagebreak

6. Both of your bar charts should have a significant number of NAs! In R, `NA` is used to indicate a missing value. The code below filters out the respondents who were `NA` for physical activity, and makes a histogram of their ages. What does this plot tell you about the missing values?

> Write your response here.

```{r}
NHANES %>% 
  filter(is.na(PhysActive)) %>%
  ggplot(aes(x=Age)) + 
  geom_histogram(bins=17, fill='lightblue', col='black') 

```

\vfill

7. Calculate the median and interquartile range of the distribution of the variable `Poverty` using the `favstats` function. Write a sentence explaining the median in the context of these data.

> Write your response here.


\vfill

\pagebreak

8. Compare the distribution of `Poverty` across each group in `Education` among adults (defined as individuals 25 years of age or older). Describe any trends or interesting observations.

> Write your response here.


\vfill

9. Construct a two-way table, with `PhysActive` as the row variable and `Diabetes` as the column variable. Among participants who are not physically active, what proportion have diabetes? What proportion of physically active participants have diabetes?

> Write your response here.


\vfill

