---
title: 'Lab 14: Analysis of Variance'
subtitle: 'MTH 361: Probability and Statistics in the Health Sciences'
date: "Last updated: `r Sys.Date()`"
output: 
  pdf_document
header-includes:
- |
  ```{=latex}
  \usepackage[default]{sourcesanspro}
  \usepackage[T1]{fontenc}
  ```
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, fig.width=6, fig.height=3)
```

# Review: Section 5.5 (Analysis of Variance)

\pagebreak

# Visualizing the ANOVA model

Suppose we have three groups: a control, treatment A, and treatment B. We are interested in how each treatment affects the response variable, $Y$. Our model is:

$$Y_{ij} = \mu + \alpha_i + \epsilon_{ij}$$

where $\epsilon_{ij} \sim Normal(0, \sigma)$, $\sigma$ representing the population standard deviation.

What does this model "look like"? Let...

- $\mu=10$
- $\alpha_{1} = 0$ (Control)
- $\alpha_{2} = 2$ (Treatment A)
- $\alpha_{3} = -2$ (Treatment B)

```{r, echo=FALSE}
library(tidyverse)
N <- 100
sigma <- 1
alpha <- c(0, 2, -2)
mu <- 10

x <- c(rep('Control', N), rep('Trt A', N), rep('Trt B', N))
y <- rnorm(length(x), 
           mean=mu+ifelse(x=='Control', alpha[1], 
                          ifelse(x=='Trt A', alpha[2], alpha[3])), 
           sd=sigma)
  
df <- tibble(x=x, 
             y=y, 
             alpha=c(rep(1, 100), rep(2, 100), rep(3, 100)),
             group=c(rep('Control', 100), rep('Trt A', 100), rep('Trt B', 100)))


x0 <- seq(from=8, to=12, length=100)
y0 <- dnorm(x0, mu + alpha[1], sigma)
path1 <- tibble(x=1+y0, y=x0)

x0 <- seq(from=10, to=14, length=100)
y0 <- dnorm(x0, mu + alpha[2], sigma)
path2 <- tibble(x=2+y0, y=x0)

x0 <- seq(from=6, to=10, length=100)
y0 <- dnorm(x0, mu + alpha[3], sigma)
path3 <- tibble(x=3+y0, y=x0)

library(colorblindr)

df %>% ggplot(aes(x=alpha, y=y)) + 
  geom_boxplot(aes(fill=group), alpha=0.5) + 
  scale_fill_OkabeIto() + 
  geom_path(aes(x, y), data=path1, col='#E69F00') + 
  geom_path(aes(x, y), data=path2, col='#56B4E9') + 
  geom_path(aes(x, y), data=path3, col='#009E73')
```


\pagebreak

# Practice problems

## Question 1: Is change in non-dominant arm strength after resistance training associated with genotype?

In the Functional polymorphisms Associated with Human Muscle Size and Strength study (FAMuSS), researchers examined the relationship between muscle strength and genotype at a particular location on the *ACTN3* gene. The `famuss` dataset in the `oibiostat` package contains a subset of data from the study. 

The percent change in non-dominant arm strength, comparing strength after resistance training to before training, is stored as `ndrm.ch`. There are three possible genotypes (CC, CT, TT) at the *r577x* position on the *ACTN3* gene; genotype is stored as `actn3.r577x`.

```{r}
library(oibiostat)
data(famuss)
head(famuss)
```

a) The boxplot below shows the association between change in non-dominant arm strength and *ACTN3* genotype. Describe what you see.

\vfill

```{r}
library(tidyverse)
library(colorblindr)
famuss %>%
  ggplot(aes(x=ndrm.ch)) + 
  geom_boxplot(aes(fill=actn3.r577x)) + 
  scale_fill_OkabeIto()
```

\pagebreak

b) Assess whether the assumptions for conducting an ANOVA are reasonably satisfied: 

    1. Observations are independent within and across groups, 
    2. The data within each group are nearly normal, and 
    3. The variability across the groups is about equal.
    
```{r}
library(mosaic)
favstats(~ndrm.ch|actn3.r577x, data=famuss)
```

\vfill

b) Conduct a hypothesis test to address the question of interest. Let $\alpha = 0.05$.

    1. Let the parameters $\mu_{CC}$, $\mu_{CT}$, and $\mu_{TT}$ represent the population mean change in non-dominant arm strength for individuals of the corresponding genotype. State the null and alternative hypotheses.
    
    2. Use `aov()` to fit the ANOVA model, and examine it using the `summary()` function. Identify the $F$-statistic and $p$-value. Interpret the $p$-value.

\vfill

```{r}
model <- aov(ndrm.ch ~ actn3.r577x, data=famuss)
summary(model)
```

\pagebreak

## Question 2: Tukey's honest significant differences

We know that at least one genotype location has a different mean non-dominant arm strength, but which one?

One option is to use __pairwise tests__: compare each pair of groups individually.

a) In the `famuss` study, how many pairs of comparisons would we need to do?

\vfill

b) How many pairs of comparisons would we need to do if we had...

    1. Four groups?

    \vfill

    2. Five groups?

    \vfill

    3. Six groups?
    
    \vfill
    
The more groups we're considering, the more tests we have to do!

__Multiple testing__ leads to an even bigger problem. The more tests we do, the more the Type I error rate gets inflated.

c) Suppose we need to do three hypothesis tests on a data set. Each one has a Type I error rate of $\alpha=0.05$. What is the probability that _at least one test_ results in an error?

\vfill

d) What happens if we need to do...

    1. Four tests?

    \vfill

    2. Five tests?

    \vfill

    3. Six tests?

    \vfill

__Familywise error rate__: 

\vfill

\pagebreak

As a solution, we "adjust" our p-values to control the overall familywise error rate at $\alpha=0.05$. One method for doing so is Tukey's honest significant differences.

__Tukey's "honest significant differences" (HSD)__: a method of making pairwise comparisons that adjusts critical values (t*) to control the overall familywise error rate

- Can be used to determine which pairwise comparisons in the ANOVA are statistically significant, and which ones are not
- Also referred to as "Tukey-Kramer" method
- There are lots of other adjustment methods, some are more "conservative" (like the Bonferroni adjustment mentioned in the textbook) and some are less "conservative"

To apply Tukey's HSD to an analysis of variance, use the `TukeyHSD()` function.

e) Interpret the table and confidence interval plots. Which groups are "different", and which are not?

\vfill

```{r}
TukeyHSD(model)

plot(TukeyHSD(model))
```

\pagebreak

## Question 3: BMI and educational attainment

This question uses data from the National Health and Nutrition Examination Survey (NHANES), a survey conducted annually by the US Centers for Disease Control (CDC). The dataset `nhanes.samp.adult.500 contains data for 500 participants ages 21 years or older that were randomly sampled from the complete NHANES dataset that contains 10,000 observations.

```{r}
library(NHANES)
data(nhanes.samp.adult.500)
```

The variable `BMI` contains BMI information for the study participants. The variable `Education` records the highest level of education obtained: 8$^{th}$ grade, 9$^{th}$ - 11$^{th}$ grade, high school, some college, or college degree.

a) The boxplot below shows the association between BMI and educational level. Describe what you see.

\vfill

```{r}
nhanes.samp.adult.500 %>%
  ggplot(aes(x=BMI)) + 
  geom_boxplot(aes(fill=Education)) + 
  scale_fill_OkabeIto()
```

\pagebreak

b) Examine the normality and equal variance assumptions across the groups. 

\vfill

```{r}
library(mosaic)
favstats(~BMI|Education, data=nhanes.samp.adult.500)
```

c) Use analysis of variance to address the question of interest. Let $\alpha = 0.05$. Summarize the conclusions.

\vfill

```{r}
model <- aov(BMI ~ Education, data=nhanes.samp.adult.500)
summary(model)
```

d) Would it make sense to use Tukey's HSD here? Explain why or why not.

\vfill

\pagebreak

# Problem(s) to submit

On your own, complete the following problems. You may use your notes, this lab activity, or any other resources. Submit your responses in BlueLine when you are finished.

## Self-care for dialysis patients

Researchers studying patients with kidney disease wanted to assess home self-care practices. Patients with kidney disease doing home dialysis treatments are especially susceptible to infections, so rigorous cleaning and self-care practices are especially important. They collected data on oral plaque levels from 90 subjects who were (1) doing home dialysis for less than a year, (2) doing home dialysis for 1 to 3 years, and (3) dialysis for more than three years. A higher value indicates a greater amount of plaque.

a) Use the boxplots provided to compare and contrast the plaque levels for the three groups. Do any groups show outliers or extreme skewness?

```{r, echo=FALSE}
#plaque <- read.csv("~/OneDrive - Creighton University/MTH 361 - Biostatistics/Worksheets and Labs/plaque.csv")
plaque <- read.csv("C:/Users/ads67836/OneDrive - Creighton University/MTH 361 - Biostatistics/Worksheets and Labs/plaque.csv")
plaque$GROUP <- as.factor(plaque$GROUP)
```

b) Use analysis of variance to detemrine whether there is a difference in plaque levels. Include your null and alternative hypothesis, test statistic, p-value, and conclusion.

c) Use Tukey's HSD to determine which pairs of groups (if any) are different.

d) Do you have any hesitations about this analysis?

R code on the next page.

\pagebreak

```{r}
head(plaque)

favstats(~PLAQUE|GROUP, data=plaque)

plaque %>% ggplot(aes(x=PLAQUE)) + 
  geom_boxplot(aes(fill=GROUP)) +
  labs(fill="Group", x="Plaque levels") + 
  scale_fill_OkabeIto()

model <- aov(PLAQUE ~ GROUP, data=plaque)

summary(model)
```

\pagebreak

```{r}
TukeyHSD(model)

plot(TukeyHSD(model))
```